Differential expression analysis of reprocessed AMPAD RNASeq data (All 7 brain regions with CQN normlaisation and BIC based model selection)

### 2017-07-06 code/Rmd/ALL_REPROCESSED.Rmd
Minor modifications

all.diff.exp = c(all.diff.exp, list(Diagnosis.SEX = DE))
all.fit = c(all.fit, list(Diagnosis.SEX = FIT))
```

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp3, fig.height=7, fig.width=24, eval = F}
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16, eval = F}
diffexp.genes = DE %>%
  dlply(.(Comparison, Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synQuery('select * from file where parentId == "syn3240583"')
ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }, geneset.files) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

@@

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'AMP_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('AMP_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = all.used.ids, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)

@@

# Store design matrix
DESIGN$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'AMP_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('AMP_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'AMP_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('AMP_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'AMP_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('AMP_DiffExpression.tsv', 
                name = 'Differential Expression Results (BrainRegion.Diagnosis)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

>>>>

all.diff.exp = c(all.diff.exp, list(Diagnosis.Gender = DE))
all.fit = c(all.fit, list(Diagnosis.Gender = FIT))
```

@@

Genes that are differentially expressed at an FDR <= 0.05 and folde change of 1.2 are
```{r diff.exp4.1, fig.height=10, fig.width=15}
# Cell type markers
CELL.FRAC = downloadFile('syn10163089')
all.used.ids = c(all.used.ids, 'syn10163089')

# Merge covariates to cell type fraction
NEW.COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID') %>%
  left_join(CELL.FRAC)

my.theme = theme(legend.position = 'top', axis.text.x = element_text(angle = 45, hjust = 1))
pl = list()
pl[[1]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = astrocytes, color = Diagnosis)) + geom_boxplot() + ggtitle('Astrocytes') + my.theme
pl[[2]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = endothelial, color = Diagnosis)) + geom_boxplot() + ggtitle('Endothelial') + my.theme
pl[[3]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = microglia, color = Diagnosis)) + geom_boxplot() + ggtitle('Microglia') + my.theme
pl[[4]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = neurons, color = Diagnosis)) + geom_boxplot() + ggtitle('Neurons') + my.theme
pl[[5]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = myeloid, color = Diagnosis)) + geom_boxplot() + ggtitle('Myeloid') + my.theme
pl[[6]] = ggplot(NEW.COVARIATES, aes(x = BrainRegion, y = oligodendrocytes, color = Diagnosis)) + geom_boxplot() + ggtitle('Oligodendrocytes') + my.theme
multiplot(plotlist = pl, cols = 3)

# Get design matrix
DESIGN = getDesignMatrix(NEW.COVARIATES[, c(adjust.covars, "astrocytes", "endothelial", "microglia", "myeloid", "neurons", "oligodendrocytes", "OPC"), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Fit contrast
cntr = c()
for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  cnt = c(paste0('BrainRegion.Diagnosis', region, '_AD-BrainRegion.Diagnosis', region, '_CONTROL'))
  cntr = c(cntr, cnt)
}
contrast = makeContrasts(contrasts=cntr,
                         levels = colnames(FIT$coefficients))
cnames = c()
for (region in  c('DLPFC', 'FP', 'STG', 'PHG', 'IFG', 'CER', 'TCX')){
  cnt = c(paste0(region, '_AD_CELLFRAC-', region, '_CONTROL_CELLFRAC'))
  cnames = c(cnames, cnt)
}
colnames(contrast) = cnames

all.diff.exp = c(all.diff.exp, list(Diagnosis.CellFraction = DE))
all.fit = c(all.fit, list(Diagnosis.CellFraction = FIT))
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
# Perform gene set enrichment analysis using wilcox test from limma
diffexp.genes = all.diff.exp %>%
  data.table::rbindlist(idcol='Model', use.names = T, fill = T) %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, labels = c('-1' = 'DOWN', '1' = 'UP')))

geneSet.test = diffexp.genes %>%
  plyr::dlply(.(Model, Comparison, Direction), .fun = function(x){
    unique(x$hgnc_symbol[x$adj.P.Val <= 0.05])
  })

statistics.test = diffexp.genes %>%
  plyr::dlply(.(Model, Comparison), .fun = function(x){
    tmp = dplyr::select(x, hgnc_symbol, adj.P.Val) %>% dplyr::group_by(hgnc_symbol) %>% dplyr::top_n(1,-adj.P.Val) %>% unique() %>% as.data.frame()
    rownames(tmp) = tmp$hgnc_symbol
    return(tmp)
  })
names(statistics.test) = paste(names(statistics.test),'UP',sep = '.')
statistics.test1 = statistics.test
names(statistics.test1) = gsub('.UP','.DOWN', names(statistics.test1))
statistics.test = c(statistics.test, statistics.test1)

backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
load(synGet('syn4867851')@filePath)
all.used.ids = c(all.used.ids, 'syn4867851')
geneSet.ref = GeneSets[c('KEGG_2016', 'GO_Biological_Process_2015', 'Reactome_2016', 'WikiPathways_2016')] %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = mapply(function(gs.test, st.test, gs.ref){
  plyr::ldply(gs.ref, .fun = function(y, gs.test, st.test){
    tmp = CovariateAnalysis::fisherEnrichment(gs.test, y, st.test$hgnc_symbol)
    tmp$wilcox.pval = limma::geneSetTest(which(st.test$hgnc_symbol %in% y), st.test[,'adj.P.Val'])
    return(tmp)
  }, gs.test, st.test, .parallel = T, .id = 'GeneSetName') %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'),
                  wilcox.fdr = p.adjust(wilcox.pval, method = 'fdr'))
}, geneSet.test, statistics.test[names(geneSet.test)], 
MoreArgs = list(geneSet.ref),
SIMPLIFY = F) %>%
  data.table::rbindlist(idcol = 'Comparison', use.names = T, fill = T)
```

@@

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'AMP_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('AMP_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
annotations(COV_OBJ) = annotations(CODE)
annotations(COV_OBJ)$dataSubType = 'covariates'
COV_OBJ = synStore(COV_OBJ, used = all.used.ids, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synChangeEntityACL(COV_OBJ@properties$id, new.racl)

@@

# Store design matrix
DESIGN$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'AMP_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('AMP_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
annotations(DM_OBJ) = annotations(CODE)
annotations(DM_OBJ)$dataSubType = 'designMatrix'
DM_OBJ = synStore(DM_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
synChangeEntityACL(DM_OBJ@properties$id, new.racl)

# Store differential expression results
write.table(diffexp.genes, file = 'AMP_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('AMP_DiffExpression.tsv', 
                name = 'Differential Expression Results', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store enrichment results
write.table(enrichment.results, file = 'AMP_EnrichmentResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('AMP_EnrichmentResults.tsv', 
                  name = 'GeneSet Enrichment Results', 
                  parentId = CODE$properties$id)
annotations(ENRICH_OBJ) = annotations(CODE)
annotations(ENRICH_OBJ)$dataSubType = 'enrichment'
ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store genesets
save(list = 'geneSet.test', file = 'AMP_DiffExpGeneSet.RData')
DEG_OBJ = File('AMP_DiffExpGeneSet.RData',
               name = 'Differential Expression Gene Set', 
               parentId = CODE$properties$id)
annotations(DEG_OBJ) = annotations(CODE)
annotations(DEG_OBJ)$dataSubType = 'diffExpGS'
DEG_OBJ = synStore(DEG_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

synDeleteEntityACL(CODE@properties$id)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ----------------------------   |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Results                         |  `r ENRICH_OBJ$properties$id`    |
|  Differentially Expressed Gene Sets         |  `r DEG_OBJ$properties$id`       |
